A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-20 tex/body.tex
commented out junk
added some stuff

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.95]{helvet}
\usepackage{booktabs}
\usepackage{bussproofs}
\usepackage{calc}
\usepackage{cite}
\usepackage{mathptmx}
%\usepackage{txfonts}
%\usepackage{mathrsfs}
%\usepackage{pifont}
%\usepackage{smallcap}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{subfigure}
\usepackage[usenames]{color}
%\usepackage{graphicx}
%\usepackage{newcent}
\usepackage{textcomp}
%\usepackage{tipa}
\usepackage{units}
\usepackage{url}
\usepackage[all]{xy}

@@

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcalx{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcalx{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\Terms(\Ec)}
\newcommand{\tcEc}{\Terms^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\ror}{\parallel}
%\newcommand{\ror}{{\&}}
%\newcommand{\ROR}{\kern.45em \ror \kern.45em}
\newcommand\ROR{\betweenantes}
\newcommand{\tpath}[2]{\mathcalx{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcalx{P}_{ #2 \rightarrow #3 }( #1 )}

@@

The last precondition is admittedly rather technical. Its role is to ensure that
$\ValC$ remains injective modulo $\alpha$-equivalence. A simpler, less efficient
way to proceed would be to compute $\nf{\ValC [ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t ]}$ optimistically and check
injectivity afterward. Injectivity is important to satisfy disequality constraints.
For infinite codatatypes, it is always possible to find suitable values $v$. For example,
if the longest cycle in $\Terms_{\smash{\J}}^{\Varec{x}}(\ValC)$ is of length $n$, a fresh value
can be constructed with a cycle of length $n + 1$.

@@

Prior to constructing a model for $F\!$, the theory solver does not explicitly construct $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all arguments of constructors in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes.
It simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of type $\Nondata$.
It then partitions each set in this partition in the same manner recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and $v$.

@@

However, the presence of quantified formulas poses an additional challenge in
the context of (co)datatypes. Quantified formulas may entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
When assuming that $\ty{u}$ has infinite cardinality, the calculus
presented in this paper is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete in the presence of quantified formulas.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the $\ty{stream}_{\,\ty{u}}$ codatatype %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F$.
\begin{rep}In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.\end{rep}

@@

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the datatype-specific
part of the decision procedure, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

The development version of CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured, to find
out whether these somewhat expensive rules are useful in practice. Even when the
decision procedure is disabled, some of the generated axioms typically include
basic properties of constructors and selectors, which the decision procedure
would recognize as tautologies. Z3 version 4.3.2 is included for comparison.
% when this makes sense.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}

>>>>

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.88]{helvet}
\usepackage{booktabs}
\usepackage{bussproofs}
\usepackage{calc}
\usepackage{cite}
\usepackage{mathptmx}
%\usepackage{txfonts}
%\usepackage{mathrsfs}
%\usepackage{pifont}
%\usepackage{smallcap}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{subfigure}
\usepackage[usenames]{color}
%\usepackage{graphicx}
%\usepackage{newcent}
\usepackage{textcomp}
%\usepackage{tipa}
\usepackage{units}
\usepackage{url}
\usepackage[all]{xy}

@@

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcalx{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcalx{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\Terms(\Ec)}
\newcommand{\tcEc}{\Terms^\AST(\Ec)}
\newcommand{\rn}[1]{\textsf{#1}}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\ror}{\parallel}
%\newcommand{\ror}{{\&}}
%\newcommand{\ROR}{\kern.45em \ror \kern.45em}
\newcommand\ROR{\betweenantes}
\newcommand{\tpath}[2]{\mathcalx{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcalx{P}_{ #2 \rightarrow #3 }( #1 )}

@@

The last precondition is admittedly rather technical. Its role is to ensure that
$\ValC$ remains injective modulo $\alpha$-equivalence. A simpler, less efficient
way to proceed would be to compute $\nf{\ValC [ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t ]}$ optimistically and check
injectivity afterward. Injectivity is important to satisfy disequality constraints.
For infinite codatatypes, it is always possible to find suitable values $v$.
%For example,
%if the longest cycle in $\Terms_{\smash{\J}}^{\Varec{x}}(\ValC)$ is of length $n$, a fresh value
%can be constructed with a cycle of length $n + 1$.

@@

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes.
It\begin{conf}%
\pagebreak
\noindent
\end{conf}
simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of type $\Nondata$.
It then partitions each set in this partition in the same manner recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and $v$.

@@

However, the presence of quantified formulas poses an additional challenge in
the context of (co)datatypes. Quantified formulas may entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
When assuming that $\ty{u}$ has infinite cardinality, the calculus
presented in Section~\ref{sec:the-ground-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete in the presence of quantified formulas.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the $\ty{stream}_{\,\ty{u}}$ codatatype %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$.
\begin{rep}In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.\end{rep}

@@

The efficiency properties of the procedure are mostly as described in Barrett et
al.\ \cite{barrett-et-al-2007}. In particular, it is theoretically and
empirically advantageous to delay applications of \rn{Split} as long as
possible. Similarly, \rn{Acyclic} and \rn{Unique} are fairly expensive because
they require traversing the equivalence classes, which is why they are part of
phase~2.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

The development version of CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning. Z3 4.3.2 is included for comparison.
% when this makes sense.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}
